---
title: "Sample Size Calculation: The Impact of Climate Change Interventions on Fruit Production in Belgium"
author: |
  **Project:** Multivariate and Hierarchical Data (3565)  
  Discovering Associations  
  Group 9  

output:
  pdf_document:
    latex_engine: xelatex
    extra_dependencies: ["booktabs", "float", "array"]
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
  - \usepackage{array}
---


# 1. Introduction

This report outlines the sample size calculation for the study, based on statistical hypotheses, assumptions on effect size, variance, power, significance level, and missing data. The aim is to ensure sufficient power to detect significant differences in pear quality across climate scenarios.

# 2. Sample size calculation

## Study objectives

Explore the difference between pear quality scores under the impact of 4 climate scenarios.

## Hypothesis

**Null Hypothesis ($H_0$):**
There is no significant difference in pear quality between the climate scenarios:

$$\mu_1 = \mu_2 = \mu_3 = \mu_4$$

**Alternative Hypothesis ($H_1$):**
There is at least one significant difference in pear quality score between pairs of climate scenarios:

$$\mu_i \neq \mu_j \quad \text{for} \quad i, j = 1, 2, 3, 4 \quad \text{and} \quad i \neq j$$

## Statistical modelling

$$ Y_{ki} = \beta_0 + b_k + \beta_1*Variety_k + \sum_{j=1}^{3}(\beta_{2j}*Scenario_{kj}) + \sum_{j=1}^{3}\beta_{3j}Scenario_{kj}*Variety_k + \epsilon_{ki}$$

- $Y_{ki}$: the i-th observed outcome of the k-th tree, $k = 1,2...K, i = 1,….,n_K$
- $b_k$: is the random effect of the k-th tree, $b_k \sim N(0, \sigma_{tree}^{2})$
- $\beta_{0}$: the average score of the reference group (Conference type, scenario 1)
- $Variety_k$: dummy variable (1 for Doyenne du Comice species, 0 otherwise)
- $Scenario_{kj}$: dummy variable, j = 1,2,3 for scenario 2, 3, 4 respectively
- $\epsilon_{ki}$: the residual error. $\epsilon_{kj} \sim N(0, \sigma_{res}^{2})$


## Assumptions

To calculate the required sample size, we use the following assumptions:

- **Significance Level **: \( \alpha = \frac{0.05}{6} \) (adjusted for multiple comparisons)
- **Effect Size**: 10, 15, and 20
- **Variance**: The standard deviation from the pilot study is approximately 36.23, inflated by a factor of 1.1 to \( \text{SD} = 39.85 \)
- **Power**: 80%, 85%, and 90%
- **Missing Data**: data missing rate data is expected to be 10%
- Each tree contains 32 pears, and there are $\leq6$ trees per ecotron, it is required 2 ecotrons for each scenario

<<<<<<< HEAD
## Result
  
```{r, echo=FALSE}
# Load necessary library
library(knitr)
library(kableExtra)


# Your existing code for data processing and sample size calculation
data <- read.csv("G9.pilot.data.csv")
sd <-  sd(data$quality_index)
variance_pilot <- var(data$quality_index)

# Define the inflation factor 
inflation_factor <- 1.1  

# Calculate the inflated variance
variance_inflated <- variance_pilot * inflation_factor

# Sample size calculation
alpha <- 0.05 / 6  # Adjusted alpha for multiple comparisons (Bonferroni correction)
power_levels <- c(0.80, 0.85, 0.90)  # Different power levels
effect_sizes <- c(10, 15, 20)  # Effect sizes
missing_data_percentage <- 0.10 

adjustment_factor <- 1 / (1 - missing_data_percentage)

# Sample size calculation for each effect size and power level using original and inflated variances
results <- data.frame()

for (effect_size in effect_sizes) {
  for (power in power_levels) {
    
    # Use the inflated variance
    sample_size_inflated <- power.t.test(delta = effect_size, sd = sqrt(variance_inflated), sig.level = alpha, power = power, 
                                         type = "two.sample", alternative = "two.sided")$n
    
    sample_size_adjusted <- ceiling(sample_size_inflated * adjustment_factor)
    
    # Store the results in a dataframe
    results <- rbind(results, data.frame(Effect_Size = effect_size, Power = power, 
                                        Sample_size_per_group = round(sample_size_inflated)))
  }
}

# Add row "Sample Size per Group (10% Missing Data)"
results$Sample_Size_per_group_adjusted <- ceiling(results$Sample_size_per_group * adjustment_factor)

# Rename columns
colnames(results) <- c("Effect Size", "Power", "Sample Size per Group", "Sample Size per Group (10% Missing Data)")

# Display table
results %>%
  kable("latex", caption = "Sample Size per group", position = "h!") %>%
  row_spec(0, bold = TRUE)
```

## 2.4 Experimental design

Pears per tree: 32

Number of scenario: 4 (Control, Scenario 2, Scenario 3, Scenario 4)

Number of Ecotrons: 12 (each Ecotrons one scenario)

Number of species: 2 (Conference and Doyenne du Comice pears)
=======
The sample size after adjusting for 10% missing data rate is presented in the table below.
>>>>>>> eb332b870e3b7c44ac8b3e6f5466a9fedbe256eb

```{r, echo=FALSE}
# Create data frames for all effect sizes
data_10 <- data.frame(
  Power = c(0.8, 0.85, 0.9),
  `Sample size per scenario` = c(390, 436, 496),
  `No. of trees per scenario` = c(13, 14, 16),
  `No. of ecotrons per scenario` = c(3, 3, 3),
  `No. of trees per species` = c("6-7", 7, 8),
  check.names = FALSE  
)

data_15 <- data.frame(
  Power = c(0.8, 0.85, 0.9),
  `Sample size per scenario` = c(175, 195, 222),
  `No. of trees per scenario` = c(6, 7, 7),
  `No. of ecotrons per scenario` = c(1, 2, 2),
  `No. of trees per species` = c(3, "3-4", "3-4"),
  check.names = FALSE
)

data_20 <- data.frame(
  Power = c(0.8, 0.85, 0.9),
  `Sample size per scenario` = c(100, 111, 126),
  `No. of trees per scenario` = c(4, 4, 4),
  `No. of ecotrons per scenario` = c(1, 1, 1),
  `No. of trees per species` = c(2, 2, 2),
  check.names = FALSE
)

# Transpose and combine all data
transposed_10 <- t(data_10[-1])  # Remove Power column
transposed_15 <- t(data_15[-1])
transposed_20 <- t(data_20[-1])
combined_data <- cbind(transposed_10, transposed_15, transposed_20)

# Removing dots in row name
rownames(combined_data) <- gsub("\\.", " ", rownames(combined_data))


colnames(combined_data) <- rep(c("0.8", "0.85", "0.9"), 3)

# Create table 
library(kableExtra)

kable(combined_data, format = "latex", booktabs = TRUE,
      caption = "Number of trees and ecotrons per scenario",
      align = "c",
      row.names = TRUE,
      escape = FALSE) %>%  # Important for special LaTeX characters
  add_header_above(c(" " = 1, 
                   "Effect size = 10" = 3, 
                   "Effect size = 15" = 3,
                   "Effect size = 20" = 3),
                 bold = TRUE) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"),
                font_size = 9,
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  column_spec(1, bold = TRUE, width = "5cm") %>%  # Wider first column
  column_spec(2:10, width = "1.5cm") %>%  # Adjust other columns
  row_spec(1:nrow(combined_data), extra_css = "white-space: nowrap;")
```

## Experimental design

Based on the table, it is observed that 175 pears per scenario are suggested to observe the minimal difference of 10 in the quality scores between 2 groups, at $\alpha = 0.0083$ and power $\beta = 0.8$. Hence, the experimental design is suggested as following:

- One ecotron per scenario, in total 4 ecotrons for 4 scenarios.
- 6 trees in each ecotron, with 3 trees for each pear species. To get a balance design, $32 \times 6 = 192$ pears per ecotron are required (96 pears for each species)
- $32 \times 6 \times 4 = 768$ pears is the final sample size. 

## Simulation

We also run 10,000 simulations for each sample size (from 120 to 300), with following parameters:

- Fixed effect size ($\delta = 15$) across all simulations
- Conservative variance at 1588 (inflated pilot SD by 10% to 39.85)
- Bonferroni-adjusted $\alpha = 0.0083$

Outcomes under the alternative hypothesis are generated (true effect exists), then two-sided t-test is performed for between the control group and one intervention group.

```{r, echo=FALSE}
# Parameters
Nvec <- seq(120, 300, by = 10)  # Sample sizes to test
n_simulations <- 10000            # Number of simulations per sample size
alpha <- 0.05/6                   # Bonferroni-adjusted alpha
delta <- 15                       # Effect size
sd_pilot <- 36.23                 # Pilot SD
sd_inflated <- sd_pilot * 1.1     # Conservative SD estimate
mean <- 55                        # Pilot mean

# Initialize results vector
sampleSizePower <- numeric(length(Nvec))
names(sampleSizePower) <- Nvec
pval = numeric(n_simulations)

set.seed(1234)  # For reproducibility

for (j in seq_along(Nvec)) {
  N <- Nvec[j]
  significant <- logical(n_simulations)  # Track significant results
  
  for (i in 1:n_simulations) {
    # Generate data
<<<<<<< HEAD
    control <- rnorm(N, mean = 50, sd = sd_inflated)
    tx1 <- rnorm(N, mean = 50 + delta, sd = sd_inflated)
<<<<<<< HEAD
    tx2 <- rnorm(N, mean = 50 + 10*delta, sd = sd_inflated)
    tx3 <- rnorm(N, mean = 50 + 20*delta, sd = sd_inflated)
=======
    tx2 <- rnorm(N, mean = 50 + delta, sd = sd_inflated)
    tx3 <- rnorm(N, mean = 50 + delta, sd = sd_inflated)
    
    scenario = c(rep(c(1,2,3,4), each = N))
    quality = c(control, tx1, tx2, tx3)
>>>>>>> faec16cd97dcf7b640a0f13d82a293cd9f2b09bc
=======
    control <- rnorm(N, mean = 55, sd = sd_inflated)
    tx1 <- rnorm(N, mean = 55 + delta, sd = sd_inflated)
>>>>>>> eb332b870e3b7c44ac8b3e6f5466a9fedbe256eb
    
    # Two-sided t-tests
    pval[i] <- t.test(control, tx1, mu = 0, alternative = "two.sided")$p.value
    
    significant[i] <- pval[i] < 0.05/6
  }
  
  # Calculate power for this sample size
  sampleSizePower[j] <- mean(significant)
}

```

```{r, echo=FALSE}
library(ggplot2)
library(cowplot)  
plot_data <- data.frame(
  SampleSize = Nvec,
  Power = sampleSizePower
)
# Create the plot with borders and caption
power_plot <- ggplot(plot_data, aes(x = SampleSize, y = Power)) +
  # Main plot elements
  geom_line(color = "#1f77b4", linewidth = 1.3) +
  geom_point(color = "#1f77b4", size = 3.5, shape = 21, fill = "white", stroke = 1.2) +
  
  # Reference lines
  geom_hline(
    yintercept = c(0.85, 0.9),
    linetype = c("dashed", "dotted"),
    color = c("#d62728", "#2ca02c"),
    linewidth = 0.8
  ) +
  
  # Annotations
  annotate(
    "text",
    x = max(Nvec),
    y = c(0.86, 0.91),
    label = c("85% Power", "90% Power"),
    color = c("#d62728", "#2ca02c"),
    hjust = 1,
    vjust = -0.5,
    size = 3.5
  ) +
  
  # Scales
  scale_y_continuous(
    limits = c(0.6, 1.0),
    breaks = seq(0.6, 1.0, 0.05),
    labels = scales::percent_format(accuracy = 1),
    expand = c(0, 0)
  ) +
  scale_x_continuous(
    breaks = seq(120, 300, 20),
    expand = c(0.02, 0.02)
  ) +
  
  # Labels
  labs(
    x = "Sample Size per Group",
    y = "Statistical Power",
    caption = "Figure 1. Results from simulation study (n = 10,000 iterations per sample size)"
  ) +
  
  # Theme customize
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.2),
    plot.title = element_text(face = "bold", hjust = 0.5, size = 11),
    plot.subtitle = element_text(hjust = 0.5, color = "gray30", size = 9),
    axis.title = element_text(size = 9),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    plot.caption = element_text(hjust = 0.5, size = 9, margin = margin(t = 10)),
    plot.margin = unit(c(1, 1, 1, 1), "cm")
  )

# Add title and subtitle with proper borders
final_plot <- ggdraw(power_plot) + 
  draw_label("Statistical Power Analysis", 
             x = 0.5, y = 0.97, size = 11, fontface = "bold") 

# Display the plot
final_plot
```

The simulation results indicate that with 190 observations per scenario, the observed power is 85%. This finding gives supporting evidence that the sample size that is suggested above is enough to acquire approximately 85% power.

# 4. Reference
Vandendijck, Y. (2025). Introduction to Sample Size Calculations [Lecture slides]. Multivariate and Hierarchical Data project, Hasselt University


